Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Forecast Model

### 2025-10-30 analysis/paper/access-ice.qmd
Adds Matt's comments

---
title: Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Forecast Model
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-mathphys-num
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    classoption: referee,lineno
    fig-width: 5  
    fig-pos: "!htb"
header-includes: | 
   \usepackage[section]{placeins}
author:
  - name: Elio Campitelli
    affiliations:
      - name: Monash University, Australia,
        id: 1
        department: School of Earth, Atmosphere and Environment
        # address: 1737 Cambridge Street
        # city: Cambridge
        # state: MA
        # postal-code: 02138
      - name: Monash University, Australia
        id: 2
        department: Securing Antarctica’s Environmental Future
        # address: 1 Oxford Street
        # city: Cambridge
        # state: MA
        # postal-code: 02138
    email: elio.campitelli@monash.edu
    attributes:
      corresponding: true
  - name: Ariaan Purich
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julie Arblaster
    affiliations:
      - ref: 1
      - ref: 2
  - name: Eun-Pa Lim
    affiliations:
      - department: Research, Bureau of Meteorology
        id: 3
  - name: Matthew Wheeler
    affiliations:
      - ref: 3      
  - name: Phillip Reid
    affiliations:
      - ref: 3      
keywords: [sea ice, seasonal predictability, initial conditions, forecasting]
bibliography: references.bib

@@

Accurate Antarctic sea-ice forecasts are crucial for climate monitoring and operational planning, yet they remain challenging due to model biases and complex ice-ocean-atmosphere interactions. 
The two versions of the Australian Bureau of Meteorology's ACCESS seasonal forecast system, +S1 and +S2, use identical model configuration and differ only in their initial conditions; primarily in that +S2 does not assimilate sea-ice observations, whereas +S1 does.   
This provides an opportunistic experiment to assess the role of initial conditions on Antarctic sea-ice forecasts using more than 20 years of fully coupled simulations with two 9-member ensembles.
Our analysis reveals that both systems experience an extended melt season and delayed growth phase compared with observations. 
This leads to a significant negative sea-ice extent bias, which is corrected only in +S1 by the data assimilation system. 
The impact of the differing initial conditions on forecast errors varies dramatically by season: summer and autumn initial conditions (January-April) provide predictive skill for up to three months, with February initial conditions being particularly crucial. 
In contrast, winter forecasts of the two systems are statistically indistinguishable after only two weeks. 
Regional analysis of forecast skill suggests that this winter predictability barrier is most dramatic over East Antarctica, where even +S1 shows negative skill. 
These findings highlight the critical importance of comprehensive year-round sampling in predictability studies and suggest that operational sea-ice data assimilation efforts should prioritize the summer-autumn period when initial conditions have maximum impact on forecast skill.

@@

However, progress in Antarctic sea-ice forecasting system has lagged behind Arctic sea-ice forecasts due to model biases, inherent large variability and complexity [@zampieri2019; @gao2024a].
Dynamical seasonal forecasts of summer Antarctic sea ice have been shown to perform worse than relatively simpler statistical methods [@massonnet2023] and machine learning approaches (e.g. @dong2024, @lin2025), which also underscores the need for better understanding of sea-ice dynamics, and drivers of its variability.

Good initial conditions are generally required for a good forecast, however, it is not entirely known to what extent accurate sea-ice initial conditions affect the quality of the forecast and at what timescales.
Exploring seasonal predictions of Arctic sea ice, @guemas2016 found that sea-ice initial conditions are important in autumn to predict summer sea ice, but the impact wasn’t as dramatic when predicting winter sea ice.
@day2014 also found seasonally-varying differences in the effect of initialisation, noting that accurate Arctic sea-ice thickness lead to improved sea-ice forecasts initialised in July but not when initialised in January.

For the Antarctic, @holland2013 studied the initial-value predictability of Antarctic sea ice in a perfect model study using the CCSM3 model.
They found that sea-ice and ocean initial conditions provide predictive information to forecast sea-ice edge location several months in advance and that some predictability is retained for up to two years thanks to ocean heat content anomalies that are advected eastward.
This is in contrast with @marchi2020, who ran perfect model experiments to argue that uncertainty in the predicted atmospheric state and evolution is the main driver of uncertainty in Antarctic sea-ice extent prediction on seasonal timescales, with sea-ice and ocean initial conditions having lesser importance.
More recently, @morioka2022 studied decadal forecasts of Antarctic sea ice and found that initialising ocean and sea ice improved the correlation between simulated and observed sea-ice concentration evolution in the Amundsen–Bellingshausen Sea.
It is hard to compare these studies since they are based on forecasts initialised at different times of the year and different frameworks: @holland2013 ran 20 ensemble members initialised on the 1st of January of a particular year, @marchi2020 ran forecasts from the 1st of March and 1st of September and @morioka2022 run forecasts only from the 1st of March.
@marchi2020 also used a coupled ocean–sea-ice model instead of a fully coupled model like @holland2013 did.
@morioka2022 used observed sea-ice initial conditions and compared with observations, while @marchi2020 and @holland2013 are perfect model studies.

>>>>

---
title: Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Forecast Model
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-mathphys-num
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    classoption: referee,lineno
    fig-width: 5  
    fig-pos: "!htb"
header-includes: | 
   \usepackage[section]{placeins}
author:
  - name: Elio Campitelli
    affiliations:
      - name: Monash University, Australia,
        id: 1
        department: School of Earth, Atmosphere and Environment
        # address: 1737 Cambridge Street
        # city: Cambridge
        # state: MA
        # postal-code: 02138
      - name: Monash University, Australia
        id: 2
        department: Securing Antarctica’s Environmental Future
        # address: 1 Oxford Street
        # city: Cambridge
        # state: MA
        # postal-code: 02138
    email: elio.campitelli@monash.edu
    attributes:
      corresponding: true
  - name: Ariaan Purich
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julie Arblaster
    affiliations:
      - ref: 1
      - ref: 2
  - name: Eun-Pa Lim
    affiliations:
      - department: Research, Bureau of Meteorology, Australia
        id: 3
  - name: Matthew C. Wheeler
    affiliations:
      - ref: 3      
  - name: Phillip Reid
    affiliations:
      - ref: 3      
keywords: [sea ice, seasonal predictability, initial conditions, forecasting]
bibliography: references.bib

@@

Accurate Antarctic sea-ice forecasts are crucial for climate monitoring and operational planning, yet they remain challenging due to model biases and complex ice-ocean-atmosphere interactions. 
The two versions of the Australian Bureau of Meteorology's ACCESS seasonal forecast system, +S1 and +S2, use identical model configuration and differ only in their initial conditions; primarily in that +S2 does not assimilate sea-ice observations, whereas +S1 does.   
This provides an convenient opportunistic experiment to assess the role of initial conditions on Antarctic sea-ice forecasts using more than 20 years of fully coupled simulations with two 9-member ensembles.
Our analysis reveals that both systems experience an extended melt season and delayed growth phase compared with observations. 
This leads to a significant negative sea-ice extent bias, which is corrected only in +S1 by the data assimilation system. 
The impact of the differing initial conditions on forecast errors varies dramatically by season: summer and autumn initial conditions (January-April) provide predictive skill for up to three months, with February initial conditions being particularly crucial. 
In contrast, winter forecasts of the two systems are statistically indistinguishable after only two weeks. 
Regional analysis of forecast skill suggests that this winter predictability barrier is most dramatic over East Antarctica, where even +S1 shows negative skill. 
These findings highlight the critical importance of comprehensive year-round sampling in predictability studies and suggest that operational sea-ice data assimilation efforts should prioritize the summer-autumn period when initial conditions have maximum impact on forecast skill.

@@

However, progress in Antarctic sea-ice forecasting system has lagged behind Arctic sea-ice forecasts due to model biases, and inherent large variability and complexity [@zampieri2019; @gao2024a].
Dynamical seasonal forecasts of summer Antarctic sea ice have been shown to perform worse than relatively simpler statistical methods [@massonnet2023] and machine learning approaches (e.g. @dong2024, @lin2025), which also underscores the need for better understanding and physical modelling of sea-ice dynamics, and drivers of its variability.

Good initial conditions are generally required for a good forecast, however, it is not entirely known to what extent accurate sea-ice initial conditions affect the quality of the forecast and at what timescales.
Exploring seasonal predictions of Arctic sea ice, @guemas2016 found that sea-ice initial conditions are important in autumn to predict summer sea ice, but the impact wasn’t as dramatic when predicting winter sea ice.
@day2014 also found seasonally-varying differences in the effect of initialisation, noting that accurate Arctic sea-ice thickness leads to improved sea-ice forecasts initialised in July but not when initialised in January.

For the Antarctic, @holland2013 studied the initial-value predictability of Antarctic sea ice in a perfect model study using the CCSM3 model.
They found that sea-ice and ocean initial conditions provide predictive information to forecast sea-ice edge location several months in advance and that some predictability is retained for up to two years thanks to ocean heat content anomalies that are advected eastward.
This is in contrast with @marchi2020, who ran perfect model experiments to argue that uncertainty in the predicted atmospheric state and evolution is the main driver of uncertainty in Antarctic sea-ice extent prediction on seasonal timescales, with sea-ice and ocean initial conditions having lesser importance.
More recently, @morioka2022 studied decadal forecasts of Antarctic sea ice and found that initialising ocean and sea ice improved the correlation between simulated and observed sea-ice concentration evolution in the Amundsen–Bellingshausen Sea.
It is hard to compare these studies since they are based on forecasts initialised at different times of the year and different frameworks: @holland2013 ran 20 ensemble members initialised on the 1st of January of a particular year, @marchi2020 ran forecasts from the 1st of March and 1st of September, and @morioka2022 ran forecasts only from the 1st of March.
@marchi2020 also used a coupled ocean–sea-ice model instead of a fully coupled model like @holland2013 did.
@morioka2022 used observed sea-ice initial conditions and compared with observations, while @marchi2020 and @holland2013 were perfect model studies.
